---
output: html_document
editor_options: 
  chunk_output_type: console
---
```{r setup-39, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, fig.width=10, fig.height=7) 
library(tidyverse) 
library(patchwork)  # for combining figures
library(RColorBrewer) # color palates
library(ggtree)    # for trees
library(phyloseq)
library(dendextend) # for tanglegrams
library(ggrepel)
library(ggpubr)
library(rstatix)


theme_set(theme_classic())
if (interactive()) load("tcr_objects.Rdata")
if (interactive()) load("GTEx/gtex_objects.RData")
boxdir <- file.path("figures", "boxplot_comparisons")
dir.create(boxdir, showWarnings = FALSE)
figures_path <- file.path("figures", "boxplot_comparisons")
tcr_corrs_anno <- tcr_corrs %>% 
  left_join(tcr_sample_metadata %>% select(sample_name:species), by=c("sample1" = "sample_name")) %>% 
  left_join(tcr_sample_metadata %>% select(sample_name:species), by=c("sample2" = "sample_name")) 

```

## Validating Trees

## Quantifying within vs between system distances

These figures were requested by a helpful reviewer who wished to see the between -vs within system distances quantified. For reference, the following figure shows the distances between tissues by the median GTEx summarized data:

```{r 39-figureS11-1 }
comp_df <- jsd_individuals_dist_list[["Tissues_in_TCR"]][[2]] %>% as_tibble(rownames = "rawA") %>%
  pivot_longer(cols = -rawA, names_to = "rawB") %>% 
  mutate(A=pmin(rawA, rawB), B=pmax(rawA, rawB)) %>%
  select(A, B, value) %>% distinct() %>% 
  filter(A != B ) %>%
  left_join(specific_tissue_key, by=c("A"="Tissue")) %>%
  left_join(specific_tissue_key, by=c("B"="Tissue")) %>%
  mutate(grp = ifelse(TissueSystem.x==TissueSystem.y, "Within", "Between"))

ggplot(comp_df, aes(x=TissueSystem.y, color=grp, group=interaction(TissueSystem.y, grp), y=value)) + 
  geom_boxplot(outlier.colour = NA) + geom_point(position = position_jitterdodge(jitter.height = 0, jitter.width = .1 )) +
  labs(x="Tissue System", y="JSD", title="GTEx median gene expression JSD between vs within Anatomic Systems")

```

Now, for the TCR data.  First, the per-patient boxplots.

```{r 39-figureS11-2 }

### Per patient
for (pat in unique(tcr_sample_metadata$indv_label)){#c("Pt1", "Pt2", "Pt3", "Pt5", "Pt6", "Pt7",  "Pt9")){
  print(str_glue("Processing {pat}"))
  # pat="Patient A"
  if (startsWith(pat, "Donor")) next
  pt_comp_df <- tcr_corrs_anno %>% 
    filter(indv_label.x == "Patient A" & indv_label.y == "Patient A" ) %>%
    rename(A=tissue.x, B=tissue.y, value=JSD) %>% 
    #mutate(A=pmin(tissue.x, tissue.y), B=pmax(tissue.x, tissue.y), value=JSD) %>%
    select(A, B, value) %>% distinct() %>% 
    filter(A != B) %>% 
    left_join(tissue_key %>% select(TissueTCR, TissueSystem), by=c("A"="TissueTCR")) %>%
    left_join(tissue_key %>% select(TissueTCR, TissueSystem), by=c("B"="TissueTCR")) %>% 
    mutate(grp = ifelse(TissueSystem.x==TissueSystem.y, "Within", "Between"))
  
p_ <- ggplot(pt_comp_df, aes(x=TissueSystem.y, color=grp, group=interaction(TissueSystem.y, grp), y=value)) + 
  geom_boxplot(outlier.colour = NA) + 
  geom_point(position = position_jitterdodge(jitter.height = 0, jitter.width = .1 ), alpha=.5)  +

ggplot(pt_comp_df %>% group_by(TissueSystem.y) %>% filter(n_distinct(grp) > 1), aes(x=TissueSystem.y, color=grp, group=interaction(TissueSystem.y, grp), y=value)) +
  geom_boxplot(outlier.colour = NA, alpha=.4) + 
  geom_point( alpha=.5, position = position_jitterdodge(jitter.height = 0, jitter.width = .1 )) + 
  stat_compare_means() + 
  plot_layout(nrow = 2)
ggsave(p_, filename = file.path(boxdir, str_glue("{pat}_comparisons.png")), width = 9, height = 8,dpi = 200)
}
```

Then the same was calculated across all samples:

```{r 39-figureS11-3 }


tcr_corrs_anno <- tcr_corrs %>% 
  left_join(tcr_sample_metadata %>% select(sample_name:species), by=c("sample1" = "sample_name")) %>% 
  left_join(tcr_sample_metadata %>% select(sample_name:species), by=c("sample2" = "sample_name")) 

pt_comp_df_all <- tcr_corrs_anno %>% 
  filter(indv_label.x == indv_label.y) %>% 
  mutate(A=pmin(tissue.x, tissue.y), B=pmax(tissue.x, tissue.y), value=JSD, pt=indv_label.x, org=org.x, stype=gvhd.x) %>%
  select(A, B, value, pt, org, stype) %>% distinct() %>% 
  filter(A != B) %>% 
  filter(A != "graft_blood" & B != "graft_blood") %>%
  left_join(tissue_annotations, by=c("A"="tissue")) %>%
  left_join(tissue_annotations, by=c("B"="tissue")) %>%
  mutate(grp = ifelse(between_within_groups.x==between_within_groups.y, "Within", "Between")) 

for (thisorg in c("Mouse", "Human")){
  if (thisorg == "Mouse"){
    thisdata <- pt_comp_df_all %>% 
      filter(between_within_groups.x != "") %>% 
      filter(org == thisorg) %>% 
      filter(stype == "GvHD") 
  } else{
    thisdata <- pt_comp_df_all %>% 
      filter(between_within_groups.x != "") %>% 
      filter(org == thisorg)
}
# find all the tissues with within-group comparisons
stat.test <- thisdata %>%
  group_by(between_within_groups.x, stype) %>% 
  filter(n_distinct(grp) > 1) %>%
  rstatix::kruskal_test(value ~ grp) %>%
  adjust_pvalue(method = "BH") %>%
  add_significance("p.adj",
                   cutpoints = c(0, 1e-04, 0.001, 0.01, 0.05, 1),
                   symbols = c("****", "***", "**", "*", "ns")
  ) %>%
  mutate(group1="Between",  group2="Within") %>%
  add_xy_position(x = "grp", dodge = 0.8)


p_between_within <- ggplot(thisdata, aes(x=grp, color=grp, group=grp, y=value)) + 
  geom_boxplot(outlier.colour = NA) +
  geom_point(position = position_jitterdodge(jitter.height = 0, jitter.width = .1 )) + 
  labs(x="System", y="JSD", color="Comparisons") + 
  facet_grid(stype~between_within_groups.x, scales = "free", space = "free_x") +
  scale_color_manual(values = c("#3949AB", "#64B5F6"), breaks = c("Between", "Within"), labels=c("Between systems", "Within systems")) + 
  theme(legend.position = "bottom") + 
 scale_x_discrete(position = "top") +
stat_pvalue_manual(
  stat.test, label = "{p.adj.signif}", tip.length = 0
  ) + theme(
    axis.ticks.x=element_blank(), axis.text.x = element_blank(),
    ) +
  gridExtra::tableGrob(tissue_annotations %>% select(between_within_groups,  tissue) %>% distinct() %>% arrange(desc(between_within_groups)), 
                       theme = gridExtra::ttheme_minimal( base_size=7),
                       rows=NULL) + 
  plot_layout(widths = c(8, 2))

ggsave(p_between_within, filename = file.path(boxdir, paste0("between_within_", thisorg, ".pdf")), width = 16, height = 8)
}
```


Next, we shuffled the classes for the systems with multiple tissues represented in the data (limiting to human samples for simplicity).  The probability weight of the sampling was set to mimic the comparisons at hand
```{r 39-figureS11-4 }
(weights <- pt_comp_df_all %>%
  filter(org == "Human") %>%
  group_by(plotgroup.x, org) %>%
  filter(n_distinct(grp) > 1) %>%
  mutate(plotgroup_n = n()) %>%
  group_by(plotgroup.x, org, grp) %>% 
  dplyr::summarize(pn =plotgroup_n[1], n=n(), weight=n()/pn) 
)
between_weight <- weights %>% filter(grp=="Between") %>% pull(weight) %>% mean()
```


The sampling was performed 1000 times, and plotted below.  

> To assess the stability of the comparisons within systems


```{r 39-figureS11-5 }
tissues_for_comparisons <- pt_comp_df_all %>%
  filter(org == "Human" ) %>% 
  filter(between_within_groups.x != "") %>%
  group_by(between_within_groups.x, org) %>% filter(n_distinct(grp) > 1) %>% pull(between_within_groups.x) %>% unique()

reps = 1000

pt_comp_df_all_shuff <- map(1:reps, .f=function(x){
  set.seed(x)
  pt_comp_df_all %>% 
    filter(org == "Human") %>% # stype == "GvHD") %>%
    filter(between_within_groups.x != "") %>%
    filter(between_within_groups.x %in% tissues_for_comparisons) %>% 
    group_by(between_within_groups.x, org)  %>% 
    mutate(
      shuffle=x, 
      fauxgrp=sample(c("Within", "Between"), replace = TRUE, size = n(), prob = c(1-between_weight, between_weight)))
}) %>% bind_rows()

# only plot the first 6 cause 100 is a bit excessive
p_inv_shuff <-ggplot(pt_comp_df_all_shuff %>% filter(shuffle <7), aes(x=between_within_groups.x, color=fauxgrp, group=interaction(between_within_groups.x, fauxgrp), y=value)) + geom_boxplot(outlier.colour = NA) + geom_point(position = position_jitterdodge(jitter.height = 0, jitter.width = .1 )) + stat_compare_means() + facet_wrap(~shuffle, ncol = 2) + theme(axis.text.x = element_text(angle=90, hjust=1)) + labs(caption="Each facet is a replicate of randomly assigning the 'Within'/'Between' grouping to comparisons relating to a anatomic system. \nFirst 6 replicates shown for clarity." ) 
ggsave(p_inv_shuff, filename = file.path(boxdir, "shuffles_between_within_first_six.pdf"), width = 16, height = 10)



shuff_summary <- pt_comp_df_all_shuff %>% 
  group_by(between_within_groups.x, shuffle) %>% 
  rstatix::kruskal_test(value ~ fauxgrp) %>%
  adjust_pvalue(method = "BH")

stat.test.allhuman <- pt_comp_df_all %>% 
      filter(between_within_groups.x != "") %>% 
      filter(org == thisorg) %>% 
  group_by(between_within_groups.x) %>% 
  filter(n_distinct(grp) > 1) %>%
  rstatix::kruskal_test(value ~ grp) %>%
  adjust_pvalue(method = "BH") %>%
  add_significance("p.adj",
                   cutpoints = c(0, 1e-04, 0.001, 0.01, 0.05, 1),
                   symbols = c("****", "***", "**", "*", "ns")
  ) %>%
  mutate(group1="Between",  group2="Within") %>%
  add_xy_position(x = "grp", dodge = 0.8)


perm_test <- shuff_summary %>% 
  left_join(stat.test.allhuman %>% select(between_within_groups.x, p.adj) %>% rename(true_p.adj=p.adj), by="between_within_groups.x") %>% 
  group_by(between_within_groups.x, true_p.adj) %>% 
  summarize(n=n(), n_less_than_true=sum(p <= true_p.adj)) %>%
  mutate(permuted_p_text = str_glue("{n_less_than_true} / {n}"))

(p_shuff_summary <-   ggplot(shuff_summary, aes(y=between_within_groups.x, group=between_within_groups.x, x=p)) + 
    geom_violin(alpha=.2) +
    geom_jitter(width = 0, height=.1, alpha=.2, color="black") + 
    geom_point(data=stat.test.allhuman, aes(x=p.adj, y=between_within_groups.x, color=between_within_groups.x), shape=8, size=3, stroke=2) + 
    scale_color_brewer(palette = "Dark2", guide="none") +
    geom_text(data=perm_test, aes(x=ifelse(true_p.adj > .1, true_p.adj, .1), y=between_within_groups.x, color=between_within_groups.x, label=permuted_p_text), nudge_y = -.25)  +
    labs(x="P-value", y="System",  title="System Permutations", caption=str_wrap("Asterisks show adjusted p-value of true system labels; black points points and violin outline show non-adjusted p-values of permutated system labels. Colored text shows number of tests equal to or less than the adjusted p-value for 1000 permutations.", width = 70)))


ggsave(p_shuff_summary, filename = file.path(boxdir, "shuffles_between_within_summary.pdf"), width = 5, height=5)

```


Lastly, we compared the broad categories of intestinal vs non-intestinal, rather than by tissue system:

```{R 39-figureS11-6 }
sets_of_comparisons <- list(
  "intestinal" = c("Large Intestine", "Small Intestine"),
  "GI" = c("Large Intestine","Small Intestine", "mLN", "Upper GI")
)


for (si in 1:length(sets_of_comparisons)){
  
  #comaprison set
  s = sets_of_comparisons[[si]]
  # name of comparison set
  sname = names(sets_of_comparisons)[[si]]
  
  intestinal_comparison_data <-  pt_comp_df_all %>% 
    filter(stype=="GvHD") %>% 
    filter(between_within_groups.x %in% s | between_within_groups.y %in% s) %>% 
    mutate(intestinal = ifelse(between_within_groups.x %in% s & between_within_groups.y %in% s, "Within", "Between"))
  
  stat.test.intestinal <- intestinal_comparison_data %>%
    group_by(org, ) %>% 
    rstatix::wilcox_test(value ~ intestinal) %>%
    adjust_pvalue(method = "BH") %>% 
    add_xy_position(x = "intestinal", dodge = 0.8)
  
  p_between_within_intestine <- ggplot(intestinal_comparison_data, aes(x=intestinal, color=intestinal, group=intestinal, y=value)) + 
    geom_boxplot(outlier.colour = NA) +
    geom_point(position = position_jitterdodge(jitter.height = 0, jitter.width = .1 )) + 
    labs(x="System", y="JSD", title = str_glue("JSD within {sname} tissues")) + 
    facet_grid(org~., scales = "free", space = "free_x") +
    scale_color_manual(values = c("#3949AB", "#64B5F6"), breaks = c( "Between", "Within"), labels=c(str_glue("Between {sname} tissues"), str_glue("Within {sname} tissues")), guide="none") + 
    theme(legend.position = "bottom") + 
    stat_pvalue_manual(stat.test.intestinal, label = "Wilcox adj. p {p}", tip.length = 0) + scale_y_continuous(limits=c(0, 1.1), expand=c(0, 0))
  
  #   stat_compare_means(comparisons = list(c( "non-intestinal", "intestinal")), label = "Wilcox adj. p {p.adj}", tip.length = 0, method = "wilcox.test") 
  
  
  nonintestinal_weight <- intestinal_comparison_data %>%
    group_by(org) %>%
    mutate(plotgroup_n = n()) %>%
    group_by(org, intestinal) %>% 
    dplyr::summarize(pn =plotgroup_n[1], n=n(), weight=n()/pn) %>% 
    filter(intestinal=="Between") %>% pull(weight) %>% mean()
  
  pt_comp_df_all_shuff_intestinal <- map(1:1000, .f=function(x){
    set.seed(x)
    intestinal_comparison_data %>% 
      group_by(org)  %>% 
      mutate(shuffle=x, fauxgrp=sample(c( "Between", "Within"), replace = TRUE, size = n(), prob = c(1-nonintestinal_weight, nonintestinal_weight)))
  }) %>% bind_rows()
  
  pt_comp_df_all_shuff_intestinal_summary <- pt_comp_df_all_shuff_intestinal %>% 
    group_by(org, shuffle) %>% 
    rstatix::wilcox_test(value ~ fauxgrp) %>%
    adjust_pvalue(method = "BH")
  
  this_perm_test <- pt_comp_df_all_shuff_intestinal_summary %>% 
    left_join(stat.test.intestinal  %>% select(org, p.adj) %>% rename(true_p.adj=p.adj)) %>% 
    group_by(org, true_p.adj) %>% 
    summarize(n=n(), n_less_than_true=sum(p <= true_p.adj)) %>%
    mutate(permuted_p_text = str_glue("{n_less_than_true} / {n}"))
  
  p_shuff_intestinal_summary <-   ggplot(pt_comp_df_all_shuff_intestinal_summary, aes(y=org, group=org, x=p)) + 
    geom_boxplot(outlier.colour = NA, alpha=.2) +
    geom_jitter(width = 0, height=.1, alpha=.2, color="black") + 
    scale_y_discrete(limits = rev(c( "Human", "Mouse"))) +
    geom_point(data=stat.test.intestinal , aes(x=p.adj, y=rev(org), color=org), shape=8, size=3, stroke=2) + scale_color_brewer(palette = "Set2", guide="none") +
    #geom_text(data=this_perm_test, aes(x=ifelse(true_p.adj > .1, true_p.adj, .1), y=org, color=org, label=permuted_p_text), nudge_y = -.25)  +
    geom_text(data=this_perm_test, aes(x= true_p.adj, y=rev(org), color=org, label=permuted_p_text), nudge_y = -.25, nudge_x=.001)  +
    labs(x="Adjusted P-value", y="Organism", title="Significance when shuffled", subtitle=paste0(sname, " vs non-", sname) ) +
    scale_x_log10()
  
  
  
  ggsave(p_between_within_intestine + p_shuff_intestinal_summary + plot_layout(ncol=1, heights = c(.8, .3)) , filename = file.path(boxdir, str_glue("mouse_human_between_within_{sname}.pdf")), width = 4, height = 8)
  
}
```

## Raw data comparison plots
```{r 39-figureS11-7 }
# note the lack of the `distinct()` call here; we need all the (redundant) comparisons for this figure
set.seed(123)

tcr_corrs_anno = tcr_corrs_anno %>% filter(sample1 != "PtA_autopsy_marrow_TCRB") %>% filter(sample2 != "PtA_autopsy_marrow_TCRB")

for(i in  unique(tcr_corrs_anno$indv_label.x)){
  tmp_p <- ggplot(tcr_corrs_anno %>% 
           filter(indv_label.x == indv_label.y) %>% filter(indv_label.x==i)  %>%  
             left_join(tissue_annotations, by=c("tissue.x"="tissue")) %>%
             filter(tissue.x != tissue.y) %>%
           left_join(tissue_annotations, by=c("tissue.y"="tissue")) , aes(x=tissue.x, y=JSD, shape=tissue.y, color=tissue.y)) + 
    geom_jitter(width = .2, height = 0, size=3)  + 
    scale_color_manual(values=tissue_annotations$new_col, breaks=tissue_annotations$tissue) +
    scale_shape_manual(values=tissue_annotations$shape, breaks=tissue_annotations$tissue) +
    labs(x="Tissue", shape="Tissue", subtitle=i, color="Tissue") +
    theme_classic()   + theme(axis.text.x = element_text(angle=45, hjust=1)) + 
    scale_y_continuous(limits=c(0, 1))
    ggsave(plot = tmp_p, filename = file.path(boxdir, str_glue("TCR_raw_comparisons_{i}.pdf")), width = 9, height = 6)
}

#formatting individual graph to focus on comparisons with blood (Figure R1)
  tmp_gvhd <- ggplot(tcr_corrs_anno %>% 
           filter(indv_label.x == indv_label.y) %>%
             filter(indv_label.x %in% c("Patient A", "Patient D", "Patient F", "Patient G"))  %>% 
             left_join(tissue_annotations, by=c("tissue.x"="tissue")) %>%
             filter(tissue.x != tissue.y) %>% filter(tissue.x == "autopsy_blood") %>%
             filter(tissue.y != "graft_blood") %>%
             filter(tissue.y != "autopsy_marrow")%>%
             filter(tissue.y != "spleen")%>%
             filter(tissue.y != "mLN")%>%
             filter(tissue.y != "heart")%>%
             filter(tissue.y != "kidney") %>%
             filter(tissue.y != "pretx_blood") %>%  
             left_join(read.csv("data/patient_attributes.csv", na.strings=c("n/a")) %>%
              mutate(subject=paste("Patient", subject)), by=c("indv_label.x" = "subject")) %>%
             left_join(tissue_annotations, by=c("tissue.y"="tissue"))%>%
             mutate(indv_label.x = reorder(indv_label.x, GVHD_severity)), aes(x=indv_label.x, GVHD_severity, y=JSD)) +
    geom_violin() + 
    geom_jitter(aes(color=tissue_label_fig1.y) , width = .2, height = 0, size=3)  +
    scale_color_manual(values=tissue_annotations$new_col, breaks=tissue_annotations$tissue_label_fig1) +
    scale_shape_manual(values=tissue_annotations$shape, breaks=tissue_annotations$tissue_label_fig1) +
    labs(x="GVHD Patient", shape="Tissue", color="Tissue") +
    theme_classic()   + theme(axis.text.x = element_text(angle=45, hjust=1)) + 
    scale_y_continuous(limits=c(0, 1))

tmp_gvhd

ggsave(plot=tmp_gvhd ,height=4,width=5, filename=file.path(figures_path, "GVHD_seversity_JSD.pdf"))

#mean, median, standard deviation for blood/GVHD comparisons
data.frame(tmp_gvhd$data  %>% group_by(sample1)  %>%
   summarize(sample1,
    mean_JSD=mean(JSD),
    median_clones_per_patient=median(JSD),
    sd_JSD=sd(JSD),
    ))



```
## Heatmaps


## Summarizing Mouse TCR data

```{r 39-figureS11-8 }
tcr_corrs_anno_mouse <- tcr_corrs_anno %>% 
  #filter(pair_is_unique == TRUE) %>% 
  filter(org.x=="Mouse") %>% filter(org.y=="Mouse") %>% 
  filter(indv_label.x==indv_label.y) %>% 
#  filter(gvhd.x == "GvHD" & gvhd.y == "GvHD") %>% 
  #filter(tissue.x != tissue.y) %>%
  select(JSD, tissue.x, tissue.y, gvhd.x, collection_day.x ) %>%
  #select(-sample1, -sample2, -org.x, -org.y, -indv_label.x, -indv_label.y, -samp_type.x, -samp_type.y, -exp_rep.x, -exp_rep.y, -indv_tissue_label.x, -indv_tissue_label.y, -indv_id.x, -indv_id.y  ) %>%
  group_by(across(c(-JSD))) %>% 
  mutate(
    n=n(),
    JSD_mean = mean(JSD),
    JSD_median = median(JSD),
    JSD_var = var(JSD),
  ) %>%
  select(tissue.x, tissue.y, everything()) %>% 
  ungroup() 

tcr_corrs_anno_mouse_summary <- tcr_corrs_anno_mouse %>% select(-JSD) %>% distinct()

p_mouse_variance <- ggplot(tcr_corrs_anno_mouse %>% 
                             filter(gvhd.x == "GvHD") %>% 
                             filter(tissue.x != tissue.y) %>% 
                             filter(collection_day.x==14), aes(x=reorder(interaction(tissue.x, tissue.y), desc(JSD_var)), y=JSD, color=as.character(n)))+
  scale_color_discrete() + 
  labs(color="N") +
  theme(
    legend.position = c(.31, .1),
    legend.direction = "horizontal",
    legend.background = element_rect(color="gray80", fill = NA),
    #legend.background = element_blank() ,
    axis.text.x = element_blank(),
    axis.title.x =element_blank(),
  ) + 
  geom_jitter(height = 0, alpha=.3, width = .2) + 
ggplot(tcr_corrs_anno_mouse_summary %>%                
         filter(tissue.x != tissue.y) %>% 
filter(gvhd.x == "GvHD") %>% filter(collection_day.x==14), aes(x=reorder(interaction(tissue.x, tissue.y), desc(JSD_var)), y=JSD_var))+
  theme(axis.text.x = element_text(angle=45, hjust=1)) + 
  geom_point() + 
  labs(y="Variance in JSD", x="") +
  plot_layout(nrow=2) +
  plot_annotation(title="JSD and JSD variance across mouse tissue pairs (day 14)")


ggsave(filename = file.path(boxdir, "mouse_JSD_variance.pdf"), plot = p_mouse_variance, width = 9, height=6)

mouse_summary_plist <- list()
for (gr in c("GvHD", "noGvHD")){
  for (cday in c(7, 14)){
    
    this_corr_data <-  tcr_corrs_anno_mouse_summary %>% 
      filter(gvhd.x == gr) %>% 
      filter(tissue.x != "Tcells",tissue.y != "Tcells" )
    # No days for healthy mice, take all of them
    if (gr == "GvHD"){
      this_corr_data <-  this_corr_data %>% 
        filter(collection_day.x  == cday) 
    }
    tcr_agg_mice <- this_corr_data %>%   
      rename(sample1=tissue.x, sample2=tissue.y) %>% 
      select(sample1, sample2, JSD_mean) %>% distinct() %>% 
      arrange(desc(sample1), desc(sample2)) %>% 
      pivot_wider(names_from="sample2", values_from=JSD_mean) %>%
      column_to_rownames("sample1") %>%
      as.matrix()
    # the arrange() makes it possible to convert to a proper lower right tri distance matrix
    dend_<- hclust(as.dist(tcr_agg_mice), method = "single") %>% as.dendrogram()
    samples_ordered <- hclust(as.dist(tcr_agg_mice), method = "single")$labels[c( hclust(as.dist(tcr_agg_mice), method = "single")$order)]
    
    tcr_agg_mice_ordered <- tcr_agg_mice[samples_ordered, samples_ordered]
    # get the clean, ordered sampleids
    samples_ordered_df <- left_join(data.frame(tissue=samples_ordered), tissue_annotations, by="tissue")
    if (!all(samples_ordered == samples_ordered_df$sample_name)) {stop("Error assigning clean name to ordered tissues")}
    samples_ordered_clean <- samples_ordered_df %>% pull(tissue)
    
    dendextend::labels_colors(dend_) <- data.frame(Tissue = labels(dend_)) %>%
      left_join(., tissue_key) %>% pull(col)
    
    
    
    tmp <- ape::as.phylo(dend_)
    tmp$edge.length <- tmp$edge.length*2
    tmpheight <- sort(stats::cophenetic(dend_), decreasing = TRUE)[1] 
    
    # 
    tick_pos <- seq(0, 1, .1)
    ticks=data.frame(xorig=tick_pos, x=tick_pos - (1-tmpheight),
                     lab=rev(tick_pos),  yend=rep(.3, length(tick_pos))) 
    
    
    p_tcr_all <- ggtree(tmp)  %<+% tissue_annotations + xlim(-.45, 1.25) +
      geom_tiplab(aes(color=new_col, label=tissue_label), 
                  offset = 0.01,#(1-max(tmp$edge.length)) + .01, 
                  align = FALSE, linetype = NA) +
      geom_tippoint() +
      scale_linetype(guide="none") +
      scale_color_identity() +
      geom_segment(y=0, yend=0, x=min(ticks$x), xend=max(ticks$x)) + 
      annotate(geom="text",        x=ticks$x, y=-1, label=ticks$lab )   + 
      geom_segment(data=ticks, aes(y=-yend, yend=0, x=x, xend=x )) +
      theme(legend.position = c(.2, .95)) 
    mouse_summary_plist[[gr]] <- p_tcr_all
    ggsave(p_tcr_all, filename = file.path("figures", str_glue("aggregated_mouse_tcr_{gr}_day{cday}.pdf")), width=6, height=8,device = "pdf")
    
    tcr_agg_mice_ordered[upper.tri(tcr_agg_mice_ordered)] <- NA
    diag(tcr_agg_mice_ordered) <- NA
    p_tcr_all_heat <- t(tcr_agg_mice_ordered) %>% as.data.frame() %>% rownames_to_column("sample1") %>% pivot_longer(-sample1, names_to = "sample2") %>% 
      rename(thisfill=value) %>%
      filter(!is.na(thisfill)) %>%
      filter(sample1 != sample2) %>% 
      mutate(
 #       sample1=factor(sample1, levels = samples_ordered_clean, ordered = TRUE),
#        sample2=rev(factor(sample2, levels = samples_ordered_clean, ordered = TRUE)),
        lab = round(thisfill, 2)
      ) %>% ggplot( aes(x=sample1, y=sample2, fill=thisfill)) + 
      geom_tile() + 
      coord_fixed(ratio=1,  ) + 
      geom_text(aes(label=lab), color="grey50", size=1)+
      scale_fill_viridis_c(option = "A") +
      guides(fill = guide_colourbar(direction = "horizontal")) + 
      scale_y_discrete(position = "left", expand = c(0,0), limits=rev(samples_ordered_clean)) +
#      scale_y_discrete(position = "right", expand = c(0,0), breaks=samples_ordered_clean) +
      scale_x_discrete( expand = c(0,0), limits=samples_ordered_clean) +
      labs(x="", y="", fill="Mean JSD") +
      theme(
        axis.text.x = element_text(angle=45, hjust=1, size=9),
        legend.position = c(.7, .7),
        )
    
    ggsave(p_tcr_all_heat,
           filename = file.path(boxdir, str_glue("aggregated_mouse_tcr_{gr}_day{cday}_heatmap.pdf")), width=6, height=8,device = "pdf")
    ggsave(p_tcr_all_heat + scale_fill_gradient2(midpoint = .5, limits=c(0, 1)),
           filename = file.path(boxdir, str_glue("aggregated_mouse_tcr_{gr}_day{cday}_heatmap_altcolor.pdf")), width=6, height=8,device = "pdf")
    
    
    
  }
}
```

