---
output: html_document
editor_options: 
  chunk_output_type: console
---

#  Rarefaction and predicting the full TCR diversity via Preseq

> Tsoni Peled and  Nick Waters

## References
- https://www.ncbi.nlm.nih.gov/pmc/articles/PMC4885658/
- also used here https://www.nature.com/articles/nprot.2018.021?&ref=https://githubhelp.com#:~:text=the%20amount%20of%20unique%[…]d%20lines)%20from%20observed%20UMIs
- https://cran.r-project.org/web/packages/preseqR/index.html
- http://smithlabresearch.org/software/preseq/
- https://github.com/smithlabcode/preseqR
- Tutorial: https://www.ncbi.nlm.nih.gov/pmc/articles/PMC4885658/bin/NIHMS785236-supplement-Supplemental.pdf

## Packages
We used preseqR v1.2.1 due to the documentation available for that version of the package.

```{r setup-41, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
depth_dir <- file.path("figures/depth/")
dir.create(depth_dir, showWarnings = FALSE)
library(tidyverse)
library(patchwork)
# devtools::install_version("preseqR", version = "1.2.1", repos = "http://cran.us.r-project.org")
library(preseqR)

tcr_counts_gvh <- tcr_raw %>% select(sample_name, rearrangement, templates)
theme_set(theme_classic())
# The  input here, (Tsoni writing this) is the dataframe overallTCR.hist
# which has two columns, and I believe the first column here is misnamed n.
# it should be j
# The first column is meant to be ascending sorted counts j = 1,2 ... (but not necessarily sequential)
# The second column is meant to n_j, the frequency count of j.
# "one example is the word frequencies in Shakepssear's known works,
# where n_j is the number of words appeared exactly j times.

# The input data one downloads from the Smith lab look like this; which is why
# Tsoni says the first column should be j, not n!

# j	n_j
# =======
# 1	603776
# 2	73628
# 3	14113
# 4	3691
# 5	2446
# 6	1612
# 7	1148

set.seed(123456)
# To produce the "histogram" input table, need to perform count() on the templates column
tcr_hist <- tcr_counts_gvh %>%
  group_by(sample_name) %>%
  count(templates) %>%
  rename(
    j = templates,
    n_j = n
  ) %>%
  arrange(sample_name, j) %>%
  ungroup()

tcr_hist %>% glimpse()
```

We set a fold increase target of 10x the original depth sampled, per sample (eg if a sample originally yielded 20k templates, we extrapolated library diversity at 200k templates).

```{R 41-figureS12A-addressing-questions-of-sampling-depth-1}
preseq_predictions <- "preseq_data.rds"
if (file.exists(preseq_predictions)) {
  tcr_increase_results <- readRDS(preseq_predictions)
} else {
  # NOTE: this generates lots of warnings:
  # - max count before zero is les than min required count (4), sample not sufficiently deep or duplicates removed
  # that seems to happen for very diverse libraries, where it appears to be deduplicated and preseq cant run
  #

  tcr_increase_results <- map(unique(tcr_hist$sample_name), .f = function(x) {
    # set a target depth fold change
    target_depth <- 10
    # get only this sample's data
    dat <- tcr_hist[tcr_hist$sample_name == x, ] %>%
      select(j, n_j)
    # get the actual target depth (total of each line of histograms' product)
    total_depth <- sum(dat$j * dat$n_j)
    # get the library complexity estimates
    overallTCR.preseqR <- preseqR.rfa.species.accum.curve(
      as.matrix(dat),
      ss = total_depth,
      max.extrapolation = total_depth * target_depth,
      bootstrap.times = 10
    )
    # reformat
    data.frame(overallTCR.preseqR) %>%
      mutate(
        sample_name = x,
        total = total_depth,
        target = total_depth * target_depth
      )
  }) %>% bind_rows()
  saveRDS(tcr_increase_results, file = preseq_predictions)
}
p_preseq_extrap_by_tissue <- ggplot(
  tcr_increase_results %>% group_by(sample_name) %>%
    summarize(perc_increase = max(yield.estimate) / min(yield.estimate)) %>%
    left_join(tcr_sample_metadata %>% select(sample_name, indv_label, tissue, org, gvhd)) %>%
    left_join(tissue_annotations),
  aes(x = reorder(tissue_label, anatomic_order), y = perc_increase, color = new_col)
) +
  geom_boxplot(outlier.colour = NA) +
  geom_jitter() +
  scale_color_identity() +
  facet_grid(org ~ gvhd, scales = "free_x", space = "free") +
  scale_y_continuous(limits = c(1, 10), expand = c(0, 0)) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  geom_hline(yintercept = 1) +
  labs(x = "Tissue", y = "Preseq predicted fold change of unique TCRs \nwith 10x sequencing depth")


p_preseq_select_tissues_curves <- ggplot(tcr_increase_results %>%
  left_join(tcr_sample_metadata %>% select(sample_name, indv_label, tissue, org, gvhd)) %>%
  filter(tissue %in% c("graft_blood", "duodenum", "skin", "stomatch", "descending_colon")) %>%
  left_join(tissue_annotations) %>%
  filter(org == "Human"), aes(x = sample.size, y = yield.estimate, color = indv_label)) +
  geom_line() +
  facet_wrap(gvhd + org ~ tissue) +
  geom_abline(intercept = 0, slope = 1) +
  coord_equal()





tcr_preseq_increase <- tcr_increase_results %>%
  group_by(sample_name) %>%
  summarize(perc_increase = max(yield.estimate) / min(yield.estimate)) %>%
  left_join(tcr_sample_metadata %>% select(sample_name, indv_label, tissue, org, gvhd)) %>%
  filter(!grepl("[Dd]onor", sample_name)) %>%
  filter("PtD_autopsy_marrow_TCRB" != sample_name) %>%
  mutate(gr = ifelse(grepl("blood", tissue), tissue, "non-blood")) %>%
  group_by(gr, org) %>%
  do(x = broom::tidy(summary(.$perc_increase))) %>%
  unnest(x) %>%
  arrange(org, gr) %>%
  mutate(across(.cols = minimum:maximum, round, 2)) %>%
  rename(`Tissue Type` = gr, `Species` = org)
```


## Rarefaction

```{r 41-figureS12A-addressing-questions-of-sampling-depth-2}

tmp_long <- tcr_raw %>%
  filter(sample_name %in% c("PtF_jejunum_TCRB", "PtF_pretx_blood_TCRB")) %>%
  select(sample_name, templates, rearrangement)
tmp <- tmp_long %>%
  pivot_wider(names_from = sample_name, values_from = templates, values_fill = 0) %>%
  column_to_rownames("rearrangement")
vegan::rarecurve(t(tmp), step = 10, ylab = "Rearrangements", col = data.frame(sample_name = colnames(tmp)) %>% left_join(tcr_sample_metadata %>% select(sample_name, tissue)) %>%
  left_join(tissue_annotations) %>% pull(new_col))

tmpdf <- map(seq(1, max(colSums(tmp)), 10), .f = function(x) {
  data.frame(Rearrangements = vegan::rarefy(t(tmp), sample = x, se = FALSE, )) %>%
    rownames_to_column("sample_name") %>%
    mutate(`Sample Size` = x)
}) %>%
  bind_rows() %>%
  left_join(data.frame(total = colSums(tmp), sample_name = names(colSums(tmp)))) %>%
  group_by(sample_name) %>%
  filter(`Sample Size` <= total) %>%
  mutate(lab = ifelse(`Sample Size` == max(`Sample Size`), sample_name, ""))

tmp_tcr_hist <- tmp_long %>%
  group_by(sample_name) %>%
  count(templates) %>%
  rename(
    j = templates,
    n_j = n
  ) %>%
  arrange(sample_name, j) %>%
  ungroup()

# make exemplar preseq curves out to 10x
these_preseq <- map(unique(tmp_tcr_hist$sample_name), .f = function(x) {
  target_depth <- 10
  dat <- tmp_tcr_hist[tmp_tcr_hist$sample_name == x, ] %>%
    select(j, n_j)
  # get the actual target depth (total of each line of histograms' product)
  total_depth <- sum(dat$j * dat$n_j)
  # get the library complexity estimates
  overallTCR.preseqR <- preseqR.rfa.species.accum.curve(
    as.matrix(dat),
    ss = total_depth * target_depth / 10,
    max.extrapolation = total_depth * target_depth,
    bootstrap.times = 100
  )
  # reformat
  data.frame(overallTCR.preseqR) %>%
    mutate(
      sample_name = x,
      total = total_depth,
      target = total_depth * target_depth
    )
}) %>% bind_rows()

# why is this requiring me such an increase in max.overlaps? I have no idea...
p_rare <- ggplot(tmpdf, aes(x = `Sample Size`, y = Rearrangements, color = sample_name)) +
  geom_abline(slope = 1, intercept = 0, linetype = "dotted", color = "grey20") +
  geom_line(size = .7) +
  scale_color_manual(values = c("red", "purple"), guide = "none") +
  geom_line(data = these_preseq, aes(x = sample.size, y = yield.estimate), linetype = "dashed", size = .2) +
  geom_segment(data = these_preseq %>% group_by(sample_name) %>% filter(sample.size == max(sample.size)) %>%
    left_join(tmpdf %>% group_by(sample_name) %>% filter(`Sample Size` == max(`Sample Size`))), aes(x = sample.size, y = yield.estimate, xend = sample.size, yend = Rearrangements), color = "grey") +
  scale_x_continuous(limits = c(0, max(these_preseq$sample.size) * 1.05)) +
  scale_y_continuous(limits = c(0, max(these_preseq$sample.size) * 1.05)) +
  geom_text(
    data = these_preseq %>% group_by(sample_name) %>% mutate(lab = ifelse(sample.size == max(sample.size), sample_name, "")),
    aes(x = sample.size, y = yield.estimate, label = lab), nudge_y = 10000, nudge_x = -30000
  ) +
  coord_equal() +
  labs(x = "Total Depth", y = "Unique TCRs") +
  # caption="Dotted black line indicates diagonal\nSolid lines are species accumulation curves.\nDashed lines indicate preseq extrapolations.\nGray vertical lines show predicted increase \n of unique species at 10X current sequencing depth")+
  theme(plot.caption = element_text(hjust = 0), text = element_text(size = 15))

# p_rare_inset <- ggplot(tmpdf %>% filter(`Sample Size` <= min(colSums(tmp))*1.1),
#                        aes(x=`Sample Size`, y=Rearrangements, color=sample_name)) + geom_line() +
#   scale_color_manual(values=c("red", "purple"), guide="none") +
#   scale_y_continuous(expand=c(0, 0)) +
#   scale_x_continuous(expand=c(0, 0)) +
#   geom_abline(slope = 1, intercept = 0, linetype="dotted", color="grey40") + coord_equal() + labs(x="", y="")

ggsave(p_rare,
  filename = file.path(depth_dir, paste0("exemplar_rarefaction_curves-", paste0(names(tmp), collapse = "__"), ".pdf")), width = 8, height = 8
  # p_rare + inset_element(p_rare_inset, left = .1, bottom = 0.7, right = .5, top = 1)
)
ggsave(p_rare,
  filename = file.path(depth_dir, paste0("exemplar_rarefaction_curves-", paste0(names(tmp), collapse = "__"), ".png")), width = 8, height = 8
  # p_rare + inset_element(p_rare_inset, left = .1, bottom = 0.7, right = .5, top = 1)
)


tt2 <- gridExtra::ttheme_minimal(
  core = list(fg_params = list(hjust = 1, x = 0.9)),
  rowhead = list(fg_params = list(hjust = 1, x = 0.95))
)
fold_increase_tab <- gridExtra::tableGrob(tcr_preseq_increase, theme = tt2, rows = NULL)
combined_fold_increase_plot <- p_rare + fold_increase_tab + plot_layout(nrow = 2, heights = c(.8, .1))
# ggview::ggview(combined_fold_increase_plot, device = "png", width = 7, height=9)

ggsave(combined_fold_increase_plot,
  filename = file.path(depth_dir, paste0("S12-exemplar_rarefaction_curves-", paste0(names(tmp), collapse = "__"), ".png")), width = 7, height = 9
)
ggsave(combined_fold_increase_plot,
  filename = file.path(depth_dir, paste0("S12-exemplar_rarefaction_curves-", paste0(names(tmp), collapse = "__"), ".pdf")), width = 7, height = 9
)
```    

